---
title: "Gravity dataset with zero trade flows"
author: "Mauricio Vargas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Description

Gravity dataset with zero trade flows is an edited version of the full gravity dataset: The "square" gravity dataset for all possible pairs of Countries worldwide, 1948-2006, which is used in the article: Head, K., T. Mayer and J. Ries, 2010, "The erosion of colonial linkages after independence". Journal of International Economics, 81(1):1-14 (lead article).

Dataset `gravity_no_zeros` corresponds to the dataset without zero trade flows, `gravity_zeros`, on the other hand, includes zero trade flows.

# Details

In order to have a dataset suited for all functions, a cross-sectional dataset is chosen. All incomplete rows and observations with missing trade flows are therefore excluded from the dataset.

As some of the functions in the package are capable of handling zero values in trade flows and some are not, two datasets, `gravity_zeros` and `gravity_no_zeros`, are provided. 

The original dataset downloaded from [SciencesPo](http://econ.sciences-po.fr/sites/default/files/file/tmayer/data/col_regfile09.zip)
was edited in the following way:

## Reading in the dataset

```{r read, eval = FALSE}
library(foreign)

col_regfile09 <- read.dta("col_regfile09.dta")
```

## Isolation of one year

```{r isolate, eval = F}
data06 <- col_regfile09[col_regfile09$year == 2006,]
```

## Choosing variables (select columns)

```{r choose, eval= F}
data06 <- data06[,c(2, 3, 6, 8, 12, 27, 34, 4, 5, 29)]
```

## Transforming data

### Isolation of complete cases

```{r complete-cases, eval = F}
data06 <- data06[complete.cases(data06) == TRUE,]

gravity_zeros <- data06

row.names(gravity_zeros) <- 1:length(row.names(gravity_zeros))
```

### Divide GDPs by 1,000,000 for scaling purposes

```{r scaling, eval = F}
gravity_zeros$gdp_o <- gravity_zeros$gdp_o / 1000000

gravity_zeros$gdp_d <- gravity_zeros$gdp_d / 1000000
```
